---
title: "RNAseq analysis"
author: ""
date: "`r format(Sys.time(), '%B %d , %Y')`"
output: 
    html_document:
        toc: TRUE
        toc_float: FALSE
        toc_depth: 2
        toc_collapsed: FALSE
        theme: readable
        number_sections: FALSE
        coding_folding: hide
        code_download: FALSE
editor_options: 
  chunk_output_type: console
params:
  fdr: 0.05
  fc: 1.5
  fdr2: 0.01
  fc2: 2
  min_samples: 15
---

```{r setup, include=FALSE, dev='CairoPNG'}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning=FALSE, dev="CairoPNG", cache = FALSE)

options(stringsAsFactors = F)
options(scipen = 99)
library(Cairo)
library(ggplot2)
library(dplyr)
library(patchwork)

```


## SUMMARY

This report presents analysis results of RNAseq data.

## {.unnumbered .unlisted}

```{r make_tabs}
# Function to create multiple tabs
make_tabs <- function(title, contents){
  cat(title, "\n\n")
  for (cnt in contents){
    print(cnt); 
    cat("\n\n")
    cat('$$\\\\[0.05in]$$')
    cat("\n\n")
    }
  cat("\n\n")
}
```

```{r functions}
count2dgelist <- function(
    counts.tsv=NULL, feature.cols=1:8, pattern2remove="^X|.bam$", counts=NULL, features = NULL, samples = NULL, group.col = 'sample_group'
    ){
    options(stringsAsFactors = F)
    require(edgeR)
    
    if(is.null(counts)){
        counts <- read.delim(counts.tsv)
    }
    if(!is.null(pattern2remove)){
        colnames(counts) <- gsub(pattern2remove, '', colnames(counts))
    }
    if(is.null(features)){
        features <- counts[feature.cols]
        counts <- counts[-(feature.cols)]
    }
        
        
    y0 <- DGEList(counts=counts, genes=features, remove.zeros = T, samples = samples) 
    if(!is.null(samples) & group.col %in% colnames(samples)){
        y0$samples$group <- samples[,group.col]
    }
    y0 <- calcNormFactors(y0)
    
    return(y0)
}

parse_de_rds <- function(fname, min_samples = 15, feature_length = 'gene_length'){
  de <- readRDS(fname)
  de <- lapply(de, function(ide){if('error' %in% names(ide)){return(NULL)}else{ide}})
  de[sapply(de, is.null)] <- NULL
  de <- de[sort(names(de))]

  if (length(de) == 0){return(NULL)}
  
  de_sum <- bind_rows(lapply(de, function(ide){ide$summary})) %>% 
    dplyr::relocate(test_samples, control_samples, .after = last_col()) %>% 
    mutate(output_folder = basename(output_folder))
  
  de_plots <- mapply(
    function(ide, out.prefix){
      y <- ide$y
      df <- ide$df
      plot.title <- paste(gsub('_vs_', ' vs ', out.prefix), '(reference)')
      list(
        PCA = plot_pca(
          y, 
          color = y$samples$sample_group, 
          sample.label = T, 
          plot.title = "", 
          var.genes = 500, 
          min.samples = min_samples, 
          feature.length = feature_length),
        MD = plot_MD(
          df, 
          plot.title = plot.title
        ),
        volcano = plot_volcano(
          df, 
          plot.title = plot.title
        )
      )
      
    }, de, names(de), SIMPLIFY = F
  )
  
  ## return results
  return(list(sum = de_sum, plots = de_plots))
  
}

parse_de_dir <- function(indir, ss, y0, min_samples = 15, feature_length = "gene_length"){
  flist <- list.files(file.path(indir, list.files(indir, pattern = '_vs_')), pattern = '.txt$', full.names = T)
  if (length(flist) == 0){return(NULL)}
  
  ss <- ss %>% 
    dplyr::select(id, sample_group) %>% 
    unique.data.frame()
  mid <- gsub('-', '.', ifelse(grepl('^[0-9]', ss$id), paste0('X', ss$id), ss$id))
  
  ## generate summary
  sum_txt <- list.files(indir, pattern = 'summary.txt|README|readme', full.names = T)
  if (length(sum_txt)>0){
    de_sum <- read.delim(sum_txt[1], comment.char = '#')
  }else{ # may fail for complicated set up, e.g. combining multiple groups
    de_sum <- lapply(
      flist, function(fname){
        df <- read.delim(fname)
        cmp <- gsub('.txt$', '', basename(fname))
        test_group <- gsub('_vs.*', '', cmp)
        control_group <- gsub('.*_vs_', '', cmp)
        ksamples <- gsub('.*normalized.', '', unique(grep('normalized', colnames(df), value = T)))

        data.frame(
          output_folder = basename(dirname(fname)),
          test_group = test_group,
          control_group = control_group,
          features_tested = nrow(df),
          features_up = sum(df$is.sig %in% 1),
          features_down = sum(df$is.sig %in% -1), 
          FDR_cutoff = params$fdr,
          FC_cutoff = params$fc,
          features_up2 = sum(df$is.sig2 %in% 1),
          features_down2 = sum(df$is.sig2 %in% -1),
          FDR_cutoff2 = params$fdr2,
          FC_cutoff2 = params$fc2
        ) %>% 
          mutate(
            test_samples = sum(ksamples %in% gsub('-', '.',ss$id)[ss$sample_group %in% test_group]),
            control_samples = sum(ksamples %in% gsub('-', '.',ss$id)[ss$sample_group %in% control_group])
          )
      }
    ) %>% 
      bind_rows
  }
  
  
  ## make plots
  cts <- y0$counts
  genes <- y0$genes
    
  de_plots <- lapply(
    flist, function(fname){
      de <- read.delim(fname)
      kgenes <- de$gene_id
      ksamples <- gsub('.*normalized.', '', unique(grep('normalized', colnames(de), value = T)))
      y <- count2dgelist(
        counts = cts[genes$gene_id %in% kgenes, match(ksamples, gsub('-', '.', colnames(cts)))], 
        features = genes[genes$gene_id %in% kgenes, ], 
        samples = ss[match(ksamples, gsub('-', '.',ss$id)), ],
        group = 'sample_group'
      )
      
    plist <- list(
        PCA = plot_pca(
            y, 
            color = y$samples$group, 
            sample.label = T, 
            plot.title = "", 
            var.genes = 500, 
            min.samples = min_samples,
            feature.length = feature_length
            ),
        MD = plot_MD(
            de, 
            plot.title = ''
        ),
        volcano = plot_volcano(
            de,
            plot.title = ''
        )
    )
    
    return(plist)
    }
  ); names(de_plots) <- gsub('.txt$', '', basename(flist))

  ## return results
  return(list(sum = de_sum, plots = de_plots))
}

## PCA
plot_pca <- function(y = NULL, cts = NULL, features = NULL, samples = NULL, var.genes = NULL, color = NULL, plot.title = '', sample.label = T, min.samples = 15, feature.length = 'gene_length'){
    options(stringsAsFactors = F)
    require(ggrepel)
    require(edgeR)
    
  ## process data 
  if(is.null(y)){
    y <- count2dgelist(
      counts = cts, 
      features = features, 
      samples = samples,
      group.col = 'sample_group'
    )
    
  }
  
  ## compute rpkm
    if('chrom' %in% colnames(y$genes)){
        y <- y[!y$genes$chrom %in% c('chrX', 'chrY', 'chrM'), ]
    }
    
    log2rpkm <- rpkm(y, gene.length = feature.length, normalized.lib.sizes = T, log = T)
    
    if (!is.null(var.genes)){
        keep <- rank(-apply(log2rpkm, 1, var)) < var.genes
        log2rpkm <- log2rpkm[keep,]
    }
    
    ## Run PCA
    pca <- prcomp(t(log2rpkm), center = T, scale = T)
    
    ## Scree plot
    pca.variance.prop <- (pca$sdev^2)/sum(pca$sdev^2)*100
    
    ## PC1 vs PC2
    df <- cbind(pca$x[,1:2],data.frame(label=factor(rownames(pca$x))))
    if(is.null(color)){
        df$color <- y$samples$group
    }else{
        df$color <- color
    }
    
    if(length(unique(df$color))>1){
        p <- ggplot(df,aes(x=PC1,y=PC2,label=label,color=color, text=label))
    }else{
        p <- ggplot(df,aes(x=PC1,y=PC2,label=label, text=label))
    }
    p <- p +
        geom_point(shape = 1)
    
    if(sample.label & nrow(df) <= min.samples){
        p <- p +
            geom_text_repel(size = 2.4, color = 'black', position = 'jitter',max.overlaps = 80)
    }
    
    p <- p +
        labs(
            title = plot.title,
            color = '', 
            x = paste0('PC1 (',round(pca.variance.prop[1],1),'%)'),
            y = paste0('PC2 (',round(pca.variance.prop[2],1),'%)')
        )+
        theme_bw()+
        theme(
          text = element_text(size = 12),
          axis.text = element_text(size = 10),
          axis.title = element_text(size = 10),
          legend.text = element_text(size = 10),
          legend.position = 'top'
        )

    ## return data
    return(p)
    
}

## Plot MD
plot_MD <- function(df, plot.title = ""){
    require(ggplot2)
    
    p <- ggplot(df, aes(x = logCPM, y = logFC, color=factor(is.sig)))+
        geom_hline(yintercept = 0)+
        geom_point(size = 0.2)+
        scale_color_manual(
            '',
            values = c("-1" = "blue", "0" = "gray", "1" = "red"),
            breaks = c(-1,0,1), labels = c('Down', 'No Sig.', 'Up')
        )+
        labs( 
            x = "Average log CPM", 
            y = "log-fold-of-change",
            title = plot.title
        )+
        theme_bw()+
      theme(
          text = element_text(size = 12),
          axis.text = element_text(size = 10),
          axis.title = element_text(size = 10),
          legend.text = element_text(size = 10),
          legend.position = 'top'
        )

    return(p)
} 

## Plot volcano 
plot_volcano <- function(df, plot.title = ""){
    require(ggplot2)
    
    p <- ggplot(df,aes(x=logFC,y=-log10(FDR),color=factor(is.sig)))+
        geom_point(size = 0.2)+ 
        scale_color_manual(
            values = c('-1'='blue','0'='gray','1'='red'),
            breaks = c('-1','0','1'),
            labels = c('Down','No Sig.','Up'),
            drop = T
        )+
        labs(
            x='log2(fold change)',
            y='-log10(FDR)',
            color='',
            title = plot.title
        )+
        theme_bw()+
        theme(
          text = element_text(size = 12),
          axis.text = element_text(size = 10),
          axis.title = element_text(size = 10),
          legend.text = element_text(size = 10),
          legend.position = 'top'
        )

    
    return(p)
}

## recompute is.sig2
recal_sig <- function(txt, col.sig, fdr, fc){
    de <- read.delim(txt)
    de[,col.sig] <- sign(de$logFC) * (abs(de$logFC) > log2(fc)) * (de$FDR < fdr)
    write.table(de, txt, sep = '\t', quote = F, row.names = F)
    return(
        data.frame(
            source.file = basename(txt),
            modify.col = col.sig,
            features.test = nrow(de),
            features.up = sum(de[,col.sig] %in% 1),
            features.down = sum(de[,col.sig] %in% -1),
            FDR.cutoff = fdr,
            FC.cutoff = fc
        )
    )
}

## scatter plot
scatter_plot <- function(
  df, 
  xvar, xlab, xsf = 1, 
  yvar, ylab, ysf = 1, 
  color.var = 'sample_group', color.lab = '',
  text.var = 'Sample',
  size.var = NULL, size.lab = ''
){
  if (!xvar %in% colnames(df) | !yvar %in% colnames(df)){return(plot_spacer())}
  
  df <- df %>% 
    mutate(
      x = df[,xvar] * xsf, 
      y = df[,yvar] * ysf,
      color = df[,color.var],
      text = df[,text.var]
    )
  
  if (is.null(size.var)){
    p <- ggplot(df, aes(
      x = x, 
      y = y,
      color = color,
      text = text
    ))
  }else if (size.var %in% colnames(df)){
    df <- df %>% mutate(size = df[,size.var])
    p <- ggplot(df, aes(
      x = x, 
      y = y,
      color = color,
      text = text,
      size = size
    ))   
  }else{
    p <- ggplot(df, aes(
      x = x, 
      y = y,
      color = color,
      text = text
    ))    
  }
  
  p+
    geom_point(shape = 1)+
    scale_size_continuous(range = c(1,3))+
    labs( 
      x = xlab, y = ylab,
      color = color.lab, size = size.lab
    )+
    theme_bw(base_size = 10)+
    theme(
      plot.margin = unit(rep(10,4), units = 'pt'),
      text = element_text(size = 12),
      axis.text.x = element_text(hjust = 1, angle = 90),
      axis.title = element_text(size = 10),
      legend.text = element_text(size = 10)
    )
}

```

$$\\[0.5in]$$

## SAMPLE SHEET
```{r sample_sheet}
ss_csv <- 'sample_sheet.csv'
ss <- read.csv(ss_csv) %>% 
  dplyr::select(-c(single_end)) %>% 
  mutate(
    id = as.character(id),
    fastq_dir = dirname(gsub(';.*', '', fastq_1)),
    fastq_1 = basename(gsub(';.*', '', fastq_1)),
    fastq_2 = basename(gsub(';.*', '', fastq_2))
  ) %>% 
  mutate(
    fastq_dir = ifelse(grepl('illumina', fastq_dir), gsub('.*illumina/', '', fastq_dir), basename(fastq_dir))
  ) %>% 
  relocate(fastq_dir, fastq_1, fastq_2, .after = last_col()) %>% 
  relocate(id)

ss %>% 
  dplyr::rename("Sample" = "id") %>% 
  DT::datatable(
  class = 'cell-border stripe', 
  rownames = T,  
  filter = 'top', 
  editable = F, 
  extensions = 'Buttons', 
  options = list(
    pageLength = 6, 
    dom = 'Bfrtip',
    buttons = c('csv')
  ))
```



$$\\[0.5in]$$

## QC {.tabset .tabset-pills}

```{r collect_qc}
## gather multiqc ####
indir <- 'multiqc_data'
if (dir.exists(indir)){
    fname <- file.path(indir, 'multiqc_general_stats.txt')
    if(file.exists(fname)){
        stat <- read.delim(fname )%>% 
          mutate(Sample = as.character(Sample))
        colnames(stat) <- gsub('.*generalstats.', '', colnames(stat))
    }

    fname <- file.path(indir, 'multiqc_star.txt')
    if(file.exists(fname)){
        star <- read.delim(fname) %>% 
          mutate(Sample = as.character(Sample))
    }
    
    fname <- file.path(indir, 'multiqc_rna_seqc.txt')
    if(file.exists(fname)){
        rnaseqc <- read.delim(fname) %>% 
          mutate(Sample = as.character(Sample))
    }

    fname <- file.path(indir, 'rseqc_gene_body_cov.txt')
    if(file.exists(fname)){
        rseqc_gene_body_cov <- read.delim(fname) %>% 
          mutate(Sample = as.character(Sample))
    }
    
}

## collect_hs_metrics ####
indir <- 'hs_metrics'
if (dir.exists(indir)){
    flist <- list.files(indir, full.names = T)
    gatk <- data.table::rbindlist(lapply(
        flist, function(fname){
            read.delim(fname, comment.char = '#', nrows = 1) %>% 
                mutate(Sample = gsub('.hs_metrics.txt', '', basename(fname)))
        }
    ), fill = T, use.names = T)
    
}

## generate metrics summary table ####
ms <- ss %>% 
  mutate(
    fastqc_id = paste0(id, ifelse(grepl(';', fastq_1), '_merged', ''), '_R1')
    ) %>% 
    dplyr::select(id, sample_group, fastqc_id) %>% 
    dplyr::rename("Sample" ="id")

if (exists('stat')){
  ms <- ms %>% 
    left_join(
      stat %>% 
        filter(is.na(fastqc.total_sequences)) %>% 
        dplyr::select(Sample, !contains('FastQC')), 
      by = c('Sample')
    ) %>% 
    left_join(
      stat %>% 
        filter(!is.na(fastqc.total_sequences) & grepl('_R1', Sample)) %>% 
        dplyr::select(Sample, contains('fastqc.total_sequences')),
      by = c('fastqc_id' = 'Sample')
    ) %>% 
    left_join(
      stat %>% 
        filter(!is.na(fastqc_trimmed.total_sequences) & grepl('_R1', Sample)) %>% 
        dplyr::select(Sample, contains('fastqc_trimmed.total_sequences')),
      by = c('fastqc_id' = 'Sample')
    )    
  
}


if (exists('rnaseqc')){
    ms <- ms %>% left_join(rnaseqc, by = 'Sample')  
}

if ( exists('gatk')){
    ms <- ms %>% left_join(gatk, by = "Sample")
}

ms$text <- ms$Sample

```

### Input reads
```{r}
df <- ms %>% 
  mutate(
    size = 1 - fastqc_trimmed.total_sequences/fastqc.total_sequences
  ) %>% 
  mutate(
    text = paste(Sample, ':', round(size*100), '% trimmed')
  )
p <- scatter_plot(
  df, 
  xvar = 'fastqc.total_sequences', xsf = 1e-6, xlab = 'Total sequenced reads in million',
  yvar = 'fastqc_trimmed.total_sequences', ysf = 1e-6, ylab = 'Adapter-trimmed reads in million',
  size.var = 'size', text.var = 'text'
)

plotly::ggplotly(p, tooltip = 'text')
rm(p,df)
```

$$\\[0.5in]$$

### Read alignment
```{r}
df <- ms %>% 
  mutate(
    text = paste(Sample, ':', round(star.uniquely_mapped_percent), '% uniquely mapped')
  )

p <- scatter_plot(
 df,
 xvar = 'star.uniquely_mapped', xsf = 1e-6, xlab = 'Uniquely mapped reads in million',
  yvar = 'star.uniquely_mapped_percent', ysf = 1, ylab = 'Uniquely mapped reads %',
  size.var = NULL, text.var = 'text'
)
plotly::ggplotly(p, tooltip = 'text')
rm(p,df)
```

$$\\[0.5in]$$

### Gene detected and intergenic rate
```{r}
p <- scatter_plot(
 ms,
 xvar = 'Genes.Detected', xsf = 1, xlab = 'Genes detected',
  yvar = 'Intergenic.Rate', ysf = 1, ylab = 'Intergenic rate',
  size.var = NULL, text.var = 'text'
)

plotly::ggplotly(p, tooltip = 'text')
rm(p)
```

$$\\[0.5in]$$

### Exonic vs intronic rates
```{r}
p <- scatter_plot(
 ms,
 xvar = 'Exonic.Rate', xsf = 1, xlab = 'Exonic rate',
  yvar = 'Intronic.Rate', ysf = 1, ylab = 'Intronic rate',
  size.var = NULL, text.var = 'text'
)
plotly::ggplotly(p, tooltip = 'text')
rm(p)
```

$$\\[0.5in]$$

### Gene body coverage
```{r, fig.dim=c(10,6)}
fname <- file.path('multiqc_data', 'rseqc_gene_body_cov.txt')
if(file.exists(fname)){
  rseqc_gene_body_cov <- read.delim(fname) %>% 
    mutate(Sample = as.character(Sample))
  df <- rseqc_gene_body_cov %>% 
    tibble::column_to_rownames(var = 'Sample') %>% 
    as.matrix() %>% 
    reshape2::melt() %>% 
    dplyr::rename(Sample = Var1, x = Var2) %>% 
    group_by(Sample) %>% 
    mutate(
      value = value/max(value)*100,
      x = as.numeric(gsub('X','',x))
    ) %>% 
    ungroup() %>% 
    dplyr::rename('match_id' = 'Sample') %>% 
    left_join(
      ms %>% 
        dplyr::select(Sample, sample_group) %>% 
        mutate(
          match_id = ifelse(grepl('^[0-9]', Sample), paste0('V',Sample), Sample)
        ), 
      by = 'match_id'
    )
  
  p <- df %>% 
    mutate(text = Sample) %>% 
    ggplot(aes(x = x, y = value, color = sample_group, group = Sample, text = text))+
    geom_line()+
    labs(
      x = "Gene body percentile (5' -> 3')", 
      y = 'Percentile of coverage',
      color = ''
    )+
    theme_bw()

}else{
  p <- plot_spacer()
}

plotly::ggplotly(p, tooltip = 'text')
rm(p,df)
  
```


$$\\[0.5in]$$

### Metrics summary
Click a button to save as csv file. \
```{r}
if (exists('ms')){
  v <- unlist(strsplit('Sample,fastqc.total_sequences,fastqc_trimmed.total_sequences,star.uniquely_mapped,star.uniquely_mapped_percent,PCT_OFF_BAIT, ZERO_CVG_TARGETS_PCT,Genes.Detected,Exonic.Rate,Intronic.Rate,Intergenic.Rate', split = ','))
  colv <- intersect(v, colnames(ms)) 
  df <- ms[,colv]
  colnames(df) <- plyr::mapvalues(colnames(df), from = v, to = c('Sample', 'Total sequenced reads', 'Adpater-trimmed reads', 'Uniquely mapped reads', 'Unique mapping percent', 'Off-target rate', 'Zero coverage rate', 'Genes detected', 'Exonic rate', 'Intronic rate', 'Intergenic rate'))
  
  df %>%
    DT::datatable(
      class = 'cell-border stripe', 
      rownames = T, 
      filter = 'top', 
      editable = T, 
      extensions = 'Buttons', 
      options = list(
        pageLength = 10, 
        dom = 'Bfrtip',
        buttons = c('csv')
      ))
}

```

## {.unnumbered .unlisted}

$$\\[0.5in]$$


## GENE-LEVEL EXPRESSION

### PCA plot of all samples 
- PCA is based on top 500 variable genes (NOT on chr X, chr Y or chr M). \
- See the sample name by hovering over a dot. \

```{r, fig.dim=c(6,4)}
if (file.exists('gene.y0.rds')){
  y0 <- readRDS('gene.y0.rds')
}else{
  fname <- 'all_samples.gene_raw_counts.txt'
  if (file.exists(fname)){
    nsamples <- nrow(ss)
    
    gene_expr <- read.delim(fname)
    cts <- gene_expr[,tail(1:ncol(gene_expr), nsamples)]
    mid <- gsub('-', '.', ifelse(grepl('^[0-9]', ss$id), paste0('X', ss$id), ss$id))
    colnames(cts) <- ss$id[match(colnames(cts), mid)] # revert "." to "-"
    
    y0 <- count2dgelist(
    counts = cts, 
    features = gene_expr[, 1:(ncol(gene_expr)-nsamples)], 
    samples = ss[match(colnames(cts), ss$id), ],
    group = 'sample_group'
    )
    rm(gene_expr, nsamples, cts, fname)
  }
}

if(exists('y0')){
  gene.pca <- plot_pca(y0, var.genes = 500, color = y0$samples$group, sample.label = F)
}else{
  gene.pca <- plot_spacer()
}

plotly::ggplotly(gene.pca, tooltip = "text")
rm(gene.pca)

```

$$\\[0.2in]$$

### Differential gene expression
```{r}
de_rds <- 'differential_genes.rds'
de_dir <- 'differential_genes'
de <- NULL
if (file.exists(de_rds)){
  de <- parse_de_rds(de_rds, min_samples = params$min_samples)
}else if (dir.exists(de_dir)){
  de <- parse_de_dir(de_dir, ss, y0, min_samples = params$min_samples)
}
rm(de_rds, de_dir)
if(exists('y0')){rm(y0)}
```

#### DE Summary
- feature = gene; FDR = BH-corrected p-value; FC = fold change  \
- Statistics of DE genes are summarized based on two thresholds -- the first set between columns *features_up* and *FC_cutoff* and the second set between columns *features_up2* and *FC_cutoff2*. \

```{r}

if (!is.null(de)){
  de$sum %>% 
    DT::datatable(
      class = 'cell-border stripe', 
      rownames = T,  
      filter = 'top', 
      editable = F, 
      extensions = 'Buttons', 
      options = list(
        pageLength = 6, 
        dom = 'Bfrtip',
        buttons = c('csv')
      ))
}
```

$$\\[0.2in]$$

#### DE plots {.tabset .tabset-pills}

DE genes displayed in the plots below and in the delivered pdf files  were defined by the first threshold in section *DE Summary* (between columns *features_up* and *FC_cutoff*).  \

- Comparisons are named as **[test group]\_vs\_[control group]**. \
- PCA is based on top 500 variable genes (NOT on chr X, chr Y or chr M). \
\

```{r, eval=TRUE, include=TRUE, results='asis', fig.cap=c('Fig. Differential expression analysis. (A) PCA plot. (B) MA/MD plot. (C) Volcano plot'), fig.width=12, fig.height=4}
if (!is.null(de)){

  for (i in 1:length(de$plots)){
    p <- wrap_plots(de$plots[[i]], nrow = 1) + plot_annotation(tag_levels = 'A')
    
    make_tabs(
      paste("#####", names(de$plots)[i]),
      list(p)
    )
  }
  
}

```

## {.unnumbered .unlisted}

$$\\[0.5in]$$


## TRANSCRIPT-LEVEL EXPRESSION


### PCA plot of all samples 
- PCA is based on top 500 variable genes (NOT on chr X, chr Y or chr M). \
- See the sample name by hovering over a dot. \

```{r, fig.dim=c(6,4)}

if (file.exists('transcript.y0.rds')){
  y0 <- readRDS('transcript.y0.rds')
}else{
  fname <- 'all_samples.transcript_raw_counts.txt'
  if (file.exists(fname)){
    nsamples <- nrow(ss)
    
    gene_expr <- read.delim(fname)
    cts <- gene_expr[,tail(1:ncol(gene_expr), nsamples)]
    if (grepl('^counts', colnames(cts)[1])){
      colnames(cts) <- gsub('^counts.', '', colnames(cts))
      mid <- gsub('-', '.', ss$id)
    }else{
      mid <- gsub('-', '.', ifelse(grepl('^[0-9]', ss$id), paste0('X', ss$id), ss$id))
    }
    colnames(cts) <- ss$id[match(colnames(cts), mid)] # revert "." to "-"
    
    y0 <- count2dgelist(
    counts = cts, 
    features = gene_expr[, 1:(ncol(gene_expr)-nsamples)], 
    samples = ss[match(colnames(cts), ss$id), ],
    group = 'sample_group'
    )
    rm(gene_expr, nsamples, cts, fname)
  }

}

if(exists('y0')){
  transcript.pca <-plot_pca(y0, var.genes = 500, color = y0$samples$group, sample.label = F, feature.length = 'EffectiveLength')
}else{
  transcript.pca <- plot_spacer()
}

plotly::ggplotly(transcript.pca, tooltip = "text")
rm(transcript.pca)


```


$$\\[0.2in]$$

### Differential transcript expression

```{r}
de_rds <- 'differential_transcripts.rds'
de_dir <- 'differential_transcripts'
de <- NULL
if (file.exists(de_rds)){
  de <- parse_de_rds(de_rds, feature_length = "EffectiveLength", min_samples = params$min_samples)
}else if (dir.exists(de_dir)){
  de <- parse_de_dir(de_dir, ss, y0, feature_length = "EffectiveLength", min_samples = params$min_samples)
  rm(y0)
}
rm(de_rds, de_dir)
if(exists('y0')){rm(y0)}
```

#### DT Summary
- feature = transcript; FDR = BH-corrected p-value; FC = fold change  \
- Statistics of DT genes are summarized based on two thresholds -- the first set between columns *features_up* and *FC_cutoff* and the second set between columns *features_up2* and *FC_cutoff2*. \

```{r}
if (!is.null(de)){
  de$sum %>% 
    relocate(control_samples, test_samples, .after = last_col()) %>% 
    DT::datatable(
      class = 'cell-border stripe', 
      rownames = T,  
      filter = 'top', 
      editable = F, 
      extensions = 'Buttons', 
      options = list(
        pageLength = 6, 
        dom = 'Bfrtip',
        buttons = c('csv')
      ))
}
```

$$\\[0.2in]$$

#### DT plots {.tabset .tabset-pills}

Differential transcripts displayed in the plots below and in the delivered pdf files  were defined by the first threshold in section *DT Summary* (between columns *features.up* and *FC.cutoff*) .  \

- Comparisons are named as **[test group]\_vs\_[control group]**. \
- PCA is based on top 500 variable genes (NOT on chr X, chr Y or chr M). \
\

```{r, eval=TRUE, include=TRUE, results='asis', fig.cap=c('Fig. Differential expression analysis. (A) PCA plot. (B) MA/MD plot. (C) Volcano plot'), fig.width=12, fig.height=4}
if (!is.null(de)){

  for (i in 1:length(de$plots)){
    p <- wrap_plots(de$plots[[i]], nrow = 1) + plot_annotation(tag_levels = 'A')
    
    make_tabs(
      paste("#####", names(de$plots)[i]),
      list(p)
    )
  }
  
}
```


## {.unnumbered .unlisted}

$$\\[0.5in]$$

## GENE FUSIONS {.tabset .tabset-pills}
```{r}
if (dir.exists('arriba')){
  flist <- list.files('arriba', pattern = 'fusions.tsv', full.names = T)
  fusions <- lapply(
    flist,
    function(fname){
      df <- read.delim(fname)
      if (nrow(df) > 0){
        data.frame(
          id = gsub('.fusions.tsv', '', basename(fname)),
          all_fusions = nrow(df),
          low_confidence = sum(df$confidence == 'low'), 
          medium_confidence = sum(df$confidence == 'medium'),
          high_confidence = sum(df$confidence == 'high')
          )
      }else{
        data.frame(
          id = gsub('.fusions.tsv', '', basename(fname)),
          all_fusions = 0,
          low_confidence = 0, 
          medium_confidence = 0,
          high_confidence = 0
        )
      }
    }
  ) %>% 
    bind_rows() %>% 
    left_join(
      ss %>% dplyr::select(id, sample_group),
      by = 'id'
    )
}


```

### Plot
```{r}
if (exists('fusions')){
  fusions %>% 
    ggplot(aes(x = total_fusions, y = high_confidence, color = sample_group))+
    geom_point(shape = 1)
  
  p <- scatter_plot(
    fusions %>% mutate(text = id), 
    xvar = 'all_fusions', xsf = 1, xlab = 'All fusions',
    yvar = 'high_confidence', ysf = 1, ylab = 'Fusions of high confidence',
    text.var = 'text'
  )
  
plotly::ggplotly(p, tooltip = 'text')
}
```


### Metrics summary
Click a button to save as csv file. \
```{r}
if (exists('fusions')){
  fusions %>% 
    dplyr::rename("Sample" = "id") %>% 
    relocate(Sample, sample_group) %>% 
    DT::datatable(
      class = 'cell-border stripe', 
      rownames = T,  
      filter = 'top', 
      editable = F, 
      extensions = 'Buttons', 
      options = list(
        pageLength = 6, 
        dom = 'Bfrtip',
        buttons = c('csv')
      ))
}

```


## {.unnumbered .unlisted}

$$\\[0.5in]$$

## DELIVERABLES
**The following files will be delivered if relevant analyses are performed. ** \

### arriba/

Genomic alignment bam files, bam index files and detected fusions.  _work/ contains other output files by Arriba.

### {.unnumbered .unlisted}

$$\\[0.1in]$$


### differential_genes/
Differential expression at the gene level. Each comparison has a subfolder (see section *DE summary* for more details). The text file has edgeR output (logFC, logCPM, p-value, BH-adjusted FDR), normalized expression in RPKM and CPM. In column is.sig (and is.sig2) of the text file, each gene is assigned 1 for up-regulation, -1 for down-regulation and 0 for no significant difference based on FDR and log2FC cutoffs (see "**DE summary**" for cutoffs used). PCA, MD/MA and volcano plots are in the same subfolder and also included in this report under section *DE plots*.\


### {.unnumbered .unlisted}

$$\\[0.1in]$$

### differential_transcripts/
Differential expression at the transcript level. Each comparison has a subfolder (see section *DE summary* for more details). The text file has edgeR output (logFC, logCPM, p-value, BH-adjusted FDR), normalized expression in RPKM and CPM. In column is.sig (and is.sig2) of the text file, each transcript is assigned 1 for up-regulation, -1 for down-regulation and 0 for no significant difference based on FDR and log2FC cutoffs (see "**DE summary**" for cutoffs used). PCA, MD/MA and volcano plots are in the same subfolder and also included in this report under section *DE plots*.\


### {.unnumbered .unlisted}

$$\\[0.1in]$$

### expression_quantification/
*all_samples.gene_raw_counts.txt* - Raw gene-level counts of all genes in all samples. \
*all_samples.transcript_raw_counts.txt* - Raw transcript-level counts of all genes in all samples. \

### {.unnumbered .unlisted}

$$\\[0.1in]$$

### multiqc/
multiqc_report.html is a QC summary report generated using [MultiQC](https://multiqc.info). \

### {.unnumbered .unlisted}

$$\\[0.1in]$$

### QC/

QC by multiple tools. \

### {.unnumbered .unlisted}

$$\\[0.5in]$$

### salmon/
Transcript-level analysis by Salmon. \

### {.unnumbered .unlisted}

$$\\[0.1in]$$

### STAR/

Genomic alignment bam files and bam index files.  _work/ contains other output files by STAR.

## {.unnumbered .unlisted}

$$\\[0.5in]$$

## METHODS

[Fastp](https://onlinelibrary.wiley.com/doi/10.1002/imt2.107) or [Cutadapt](https://journal.embnet.org/index.php/embnetjournal/article/view/200) was applied to remove adapter sequences. [STAR](https://pubmed.ncbi.nlm.nih.gov/23104886/) with 2-pass mapping was used to align reads to the reference genome and gene annotation along with quantification of gene-level expression. [FastQC](https://www.bioinformatics.babraham.ac.uk/projects/fastqc/), [RNA-SeQC](https://pubmed.ncbi.nlm.nih.gov/22539670/), [RSeQC](https://pubmed.ncbi.nlm.nih.gov/22743226/) were used to check various QC metrics including insert fragment size, read quality, read duplication rates, rRNA rates, gene body coverage and read distribution in different genomic regions. \
Bioconductor package [edgeR](https://academic.oup.com/bioinformatics/article/26/1/139/182458) was used to detect differential expression between sample groups. Genes with low expression were excluded using edgeR function filterByExpr with min.count = 10 and min.total.count = 15. The filtered expression matrix was normalized by [TMM method](https://doi.org/10.1186/gb-2010-11-3-r25). Statistical test was using quasi-likelihood pipeline implemented in edgeR. A gene was deemed differentially expressed if meeting the cutoffs of log2 fold change and Benjamini-Hochberg adjusted p-values. \

$$\\[0.5in]$$

## ACKNOWLEDGEMENTS
(For publication, please include the following acknowledgement.)
This research was supported by the Genomics & Bioinformatics Shared Resource, RRID:SCR_022606, of the Fred Hutch/University of Washington/Seattle Children’s Cancer Consortium (P30 CA015704).

$$\\[0.5in]$$

