---
title: "RNAseq analysis"
author: ""
date: "`r format(Sys.time(), '%B %d , %Y')`"
output: 
    html_document:
        toc: TRUE
        toc_float: FALSE
        toc_depth: 2
        toc_collapsed: FALSE
        theme: readable
        number_sections: FALSE
        coding_folding: hide
        code_download: FALSE
editor_options: 
  chunk_output_type: console
params:
  dat.dir: "./"
---

```{r setup, include=FALSE, dev='CairoPNG'}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning=FALSE, dev="CairoPNG")
knitr::opts_knit$set(root.dir = params$dat.dir)

options(stringsAsFactors = F)
options(scipen = 99)
library(Cairo)
library(ggplot2)
library(dplyr)
library(patchwork)
library(kableExtra)

setwd(params$dat.dir)

```


## Summary

This document presents analysis results of RNAseq data.

$$\\[0.5in]$$

## {.unnumbered .unlisted}

```{r make_tabs}
# Function to create multiple tabs
make_tabs <- function(title, contents){
  cat(title, "\n\n")
  for (cnt in contents){
    print(cnt); 
    cat("\n\n")
    cat('$$\\\\[0.05in]$$')
    cat("\n\n")
    }
  cat("\n\n")
}
```

```{r plot_functions}
## PCA
plot_pca <- function(y, var.genes = NULL, color = NULL, plot.title = '', sample.label = T, feature.length = 'gene_length'){
    options(stringsAsFactors = F)
    require(ggrepel)
    require(edgeR)
    

    if('chrom' %in% colnames(y$genes)){
        y <- y[!y$genes$chrom %in% c('chrX', 'chrY', 'chrM'), ]
    }
    
    log2rpkm <- rpkm(y, gene.length = feature.length, normalized.lib.sizes = T, log = T)
    
    if (!is.null(var.genes)){
        keep <- rank(-apply(log2rpkm, 1, var)) < var.genes
        log2rpkm <- log2rpkm[keep,]
    }
    
    ## Run PCA
    pca <- prcomp(t(log2rpkm), center = T, scale = T)
    
    ## Scree plot
    pca.variance.prop <- (pca$sdev^2)/sum(pca$sdev^2)*100
    
    ## PC1 vs PC2
    df <- cbind(pca$x[,1:2],data.frame(label=factor(rownames(pca$x))))
    if(is.null(color)){
        df$color <- y$samples$group
    }else{
        df$color <- color
    }
    
    if(length(unique(df$color))>1){
        p <- ggplot(df,aes(x=PC1,y=PC2,label=label,color=color, text=label))
    }else{
        p <- ggplot(df,aes(x=PC1,y=PC2,label=label, text=label))
    }
    p <- p +
        geom_point(shape = 1)
    
    if(sample.label){
        p <- p +
            geom_text_repel(size = 2.4, color = 'black', position = 'jitter',max.overlaps = 80)
    }
    
    p <- p +
        labs(
            title = plot.title,
            color = '', 
            x = paste0('PC1 (',round(pca.variance.prop[1],1),'%)'),
            y = paste0('PC2 (',round(pca.variance.prop[2],1),'%)')
        )+
        theme_bw()+
        theme(
          text = element_text(size = 12),
          axis.text = element_text(size = 10),
          axis.title = element_text(size = 10),
          legend.text = element_text(size = 10),
          legend.position = 'top'
        )

    ## return data
    return(p)
    
}

## Plot MD
plot_MD <- function(df, plot.title = ""){
    require(ggplot2)
    
    p <- ggplot(df, aes(x = logCPM, y = logFC, color=factor(is.sig)))+
        geom_hline(yintercept = 0)+
        geom_point(size = 0.2)+
        scale_color_manual(
            '',
            values = c("-1" = "blue", "0" = "gray", "1" = "red"),
            breaks = c(-1,0,1), labels = c('Down', 'No Sig.', 'Up')
        )+
        labs( 
            x = "Average log CPM", 
            y = "log-fold-of-change",
            title = plot.title
        )+
        theme_bw()+
      theme(
          text = element_text(size = 12),
          axis.text = element_text(size = 10),
          axis.title = element_text(size = 10),
          legend.text = element_text(size = 10),
          legend.position = 'top'
        )

    return(p)
} 

## Plot volcano 
plot_volcano <- function(df, plot.title = ""){
    require(ggplot2)
    
    p <- ggplot(df,aes(x=logFC,y=-log10(FDR),color=factor(is.sig)))+
        geom_point(size = 0.2)+ 
        scale_color_manual(
            values = c('-1'='blue','0'='gray','1'='red'),
            breaks = c('-1','0','1'),
            labels = c('Down','No Sig.','Up'),
            drop = T
        )+
        labs(
            x='log2(fold change)',
            y='-log10(FDR)',
            color='',
            title = plot.title
        )+
        theme_bw()+
        theme(
          text = element_text(size = 12),
          axis.text = element_text(size = 10),
          axis.title = element_text(size = 10),
          legend.text = element_text(size = 10),
          legend.position = 'top'
        )

    
    return(p)
}

## recompute is.sig2
recal_sig <- function(txt, col.sig, fdr, fc){
    de <- read.delim(txt)
    de[,col.sig] <- sign(de$logFC) * (abs(de$logFC) > log2(fc)) * (de$FDR < fdr)
    write.table(de, txt, sep = '\t', quote = F, row.names = F)
    return(
        data.frame(
            source.file = basename(txt),
            modify.col = col.sig,
            features.test = nrow(de),
            features.up = sum(de[,col.sig] %in% 1),
            features.down = sum(de[,col.sig] %in% -1),
            FDR.cutoff = fdr,
            FC.cutoff = fc
        )
    )
}

## density plot
density_plot <- function(df, xvar, xlab, xsf = 1){
  require(ggplot2)
  df <- as.data.frame(df)
  if (xvar %in% colnames(df)){
    df$x <- df[,xvar] * xsf
    
    df %>% 
      ggplot(aes(x = x))+
      geom_density()+
      labs(x = xlab)+
      theme_bw()+
      theme(
        plot.margin = unit(rep(10,4), units = 'pt'),
        text = element_text(size = 12),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.position = 'none'
      )
  }else{
    return(NULL)
  }
}

## dot plot
dot_plot <- function(df, xvar, xlab, xsf = 1, yvar = 'sample_group'){
  require(ggplot2)
  df <- as.data.frame(df)
  if (xvar %in% colnames(df)){
    df$x <- df[,xvar] * xsf
    
    if (yvar %in% colnames(df)){
      df$y <- df[,yvar]
    }else{
      df$y <- ''
    }
    df %>% 
      ggplot(aes(x = x, y = y, text = text))+
      geom_jitter(width = 0, height = 0.3, shape = 1)+
      labs(x = xlab, y = '')+
      theme_bw(base_size = 10)+
      theme(
        plot.margin = unit(rep(10,4), units = 'pt'),
        text = element_text(size = 12),
        axis.text.x = element_text(hjust = 1, angle = 90),
        axis.title = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.position = 'none'
      )
  }else{
    return(NULL)
  }
}

## scatter plot
scatter_plot <- function(
  df, 
  xvar, xlab, xsf = 1, 
  yvar, ylab, ysf = 1, 
  color.var = 'sample_group', color.lab = '',
  text.var = 'Sample',
  size.var = NULL, size.lab = ''
){
  if (!xvar %in% colnames(df) | !yvar %in% colnames(df)){return(plot_spacer())}
  
  df <- df %>% 
    mutate(
      x = df[,xvar] * xsf, 
      y = df[,yvar] * ysf,
      color = df[,color.var],
      text = df[,text.var]
    )
  
  if (is.null(size.var)){
    p <- ggplot(df, aes(
      x = x, 
      y = y,
      color = color,
      text = text
    ))
  }else if (size.var %in% colnames(df)){
    df <- df %>% mutate(size = df[,size.var])
    p <- ggplot(df, aes(
      x = x, 
      y = y,
      color = color,
      text = text,
      size = size
    ))   
  }else{
    p <- ggplot(df, aes(
      x = x, 
      y = y,
      color = color,
      text = text
    ))    
  }
  
  p+
    geom_point()+
    scale_size_continuous(range = c(1,3))+
    labs( 
      x = xlab, y = ylab,
      color = color.lab, size = size.lab
    )+
    theme_bw(base_size = 10)+
    theme(
      plot.margin = unit(rep(10,4), units = 'pt'),
      text = element_text(size = 12),
      axis.text.x = element_text(hjust = 1, angle = 90),
      axis.title = element_text(size = 10),
      legend.text = element_text(size = 10)
    )
}
```

```{r read_input}
dat <- readRDS('data.rds')
ss <- dat$ss %>% 
  dplyr::select(-c(single_end)) %>% 
  mutate(
    fastq_dir = gsub('.*illumina/', '', dirname(fastq_1)),
    fastq_1 = basename(fastq_1),
    fastq_2 = basename(fastq_2)
  ) %>% 
  relocate(fastq_dir, fastq_1, fastq_2, .after = last_col()) %>% 
  relocate(id)

ms <- ss %>% 
    mutate(fastq_id = sub('.fastq.gz$', '', fastq_1)) %>% 
    dplyr::select(id, sample_group, fastq_id) %>% 
    dplyr::rename("Sample" ="id")

if ('stat' %in% names(dat)){
  ms <- ms %>% 
    left_join(
      dat$stat %>% 
      filter(is.na(FastQC_mqc.generalstats.fastqc.total_sequences)) %>% 
      dplyr::select(Sample, !contains('FastQC')), 
      by = c('Sample')
    ) %>% 
    left_join(
      dat$stat %>% 
        filter(!is.na(FastQC_mqc.generalstats.fastqc.total_sequences) & grepl('_R1_|_merged_R1', Sample)) %>% 
        dplyr::select(Sample, contains('fastqc.total_sequences')) %>% 
        mutate(Sample = Sample),
      by = c('fastq_id' = 'Sample')
    ) %>% 
    left_join(
      dat$stat %>% 
        filter(!is.na(FastQC_trimmed_mqc.generalstats.fastqc_trimmed.total_sequences) & grepl('_R1_|_merged_R1', Sample)) %>% 
        dplyr::select(Sample, contains('fastqc_trimmed.total_sequences')) %>% 
        mutate(Sample = Sample),
      by = c('fastq_id' = 'Sample')
    )    
  colnames(ms) <- gsub('.*generalstats.', '', colnames(ms))
}


if ('rnaseqc' %in% names(dat)){
  if (nrow(ms) > 0){
    ms <- ms %>% 
      left_join(dat$rnaseqc, by = 'Sample')  
  }else{
    ms <- dat$rnaseqc
  }
  
}

if ( 'gatk' %in% names(dat) ){
  if (nrow(ms) > 0){
    ms <- left_join(ms, dat$gatk, by = "Sample")
  }else{
    ms <- dat$gatk
  }
  
}

ms$text <- ms$Sample

```

```{r read_qc}
qc_plot <- dot_plot

figs <- list()

## reads ####
df <- ms %>% 
  mutate(
    size = 1 - fastqc_trimmed.total_sequences/fastqc.total_sequences
  ) %>% 
  mutate(
    text = paste(Sample, ':', round(size*100), '% trimmed')
  )
figs$reads <- scatter_plot(
  df, 
  xvar = 'fastqc.total_sequences', xsf = 1e-6, xlab = 'Total sequenced reads in million',
  yvar = 'fastqc_trimmed.total_sequences', ysf = 1e-6, ylab = 'Adapter-trimmed reads in million',
  size.var = 'size', text.var = 'text'
)

## alignment ####
df <- ms %>% 
  mutate(
    text = paste(Sample, ':', round(star.uniquely_mapped_percent), '% uniquely mapped')
  )

figs$alignment <- scatter_plot(
 df,
 xvar = 'star.uniquely_mapped', xsf = 1e-6, xlab = 'Uniquely mapped reads in million',
  yvar = 'star.uniquely_mapped_percent', ysf = 1, ylab = 'Uniquely mapped reads %',
  size.var = NULL, text.var = 'text'
)

## genes vs intergenic regions ####
figs$genes <- scatter_plot(
 ms,
 xvar = 'Genes.Detected', xsf = 1, xlab = 'Genes detected',
  yvar = 'Intergenic.Rate', ysf = 1, ylab = 'Intergenic rate',
  size.var = NULL, text.var = 'text'
)

## exonic vs intron ####
figs$exon <- scatter_plot(
 ms,
 xvar = 'Exonic.Rate', xsf = 1, xlab = 'Exonic rate',
  yvar = 'Intronic.Rate', ysf = 1, ylab = 'Intronic rate',
  size.var = NULL, text.var = 'text'
)

## off-target vs zero-coverage ####
if ('PCT_OFF_BAI' %in% colnames(ms) & 'ZERO_CVG_TARGETS_PCT' %in% colnames(ms)){
  figs$coverage <- scatter_plot(
    ms,
    xvar = 'PCT_OFF_BAIT', xsf = 1, xlab = 'Off-target rate',
    yvar = ' ZERO_CVG_TARGETS_PCT', ysf = 1, ylab = 'Zero coverage rate',
    size.var = NULL, text.var = 'text'
  )
  
}


```

```{r run_pca}
if (file.exists('gene.y0.rds')){
  y0 <- readRDS('gene.y0.rds')
  gene.pca <- plot_pca(y0, var.genes = 500, color = y0$samples$group, sample.label = F)
  if (length(unique(y0$samples$group))>30){
    y0$samples$group <- 1
    gene.pca <- plot_pca(y0, var.genes = 500, color = y0$samples$group, sample.label = F)
  }
  
}

if (file.exists('transcript.y0.rds')){
  y0 <- readRDS('transcript.y0.rds')
  transcript.pca <-plot_pca(y0, var.genes = 500, color = y0$samples$group, sample.label = F, feature.length = 'EffectiveLength')
  if (length(unique(y0$samples$group))>30){
    y0$samples$group <- 1
    ranscript.pca <-plot_pca(y0, var.genes = 500, color = y0$samples$group, sample.label = F, feature.length = 'EffectiveLength')
  }
}
```

```{r run_de}
de.rds <- 'differential_genes.rds'
if (file.exists(de.rds)){
  de <- readRDS(de.rds)
  de <- lapply(de, function(ide){if('error' %in% names(ide)){return(NULL)}else{ide}})
  de[sapply(de, is.null)] <- NULL
  
}

if ('de' %in% ls() & length(de) > 0){
  de_sum <- bind_rows(lapply(de, function(ide){ide$summary})) %>% 
    dplyr::relocate(output.folder, control.group, test.group, features.tested:FC.cutoff, control.samples:test.samples)
 
  de_plots <- mapply(
    function(ide, out.prefix){
        y <- ide$y
        df <- ide$df
        plot.title <- gsub('_vs_', ' vs ', out.prefix)
        list(
          PCA = plot_pca(
            y, 
            color = y$samples$sample_group, 
            sample.label = T, 
            plot.title = "", 
            var.genes = 500, 
            feature.length = "gene_length"),
          MD = plot_MD(
            df, 
            plot.title = ""
          ),
          volcano = plot_volcano(
            df, 
            plot.title = ""
          )
        )
        
      }, de, names(de), SIMPLIFY = F
  )
  
}
```

```{r run_dt}
dt.rds <- 'differential_transcripts.rds'
if (file.exists(dt.rds)){
  dt <- readRDS(dt.rds)
  dt <- lapply(dt, function(idt){if('error' %in% names(idt)){return(NULL)}else{idt}})
  dt[sapply(dt, is.null)] <- NULL
  
}

if ('dt' %in% ls() & length(dt) > 0){
  dt_sum <- bind_rows(lapply(dt, function(ide){ide$summary})) %>% 
    dplyr::relocate(output.folder, control.group, test.group, features.tested:FC.cutoff, control.samples:test.samples)
 
  dt_plots <- mapply(
    function(ide, out.prefix){
        y <- ide$y
        df <- ide$df
        plot.title <- gsub('_vs_', ' vs ', out.prefix)
        list(
          PCA = plot_pca(
            y, 
            color = y$samples$sample_group, 
            sample.label = T, 
            plot.title = "", 
            var.genes = 500, 
            feature.length = "EffectiveLength"),
          MD = plot_MD(
            df, 
            plot.title = ""
          ),
          volcano = plot_volcano(
            df, 
            plot.title = ""
          )
        )
        
      }, dt, names(dt), SIMPLIFY = F
  )
  
}

```

## Sample sheet
```{r sample_sheet}

ss %>% 
  DT::datatable(
  class = 'cell-border stripe', 
  rownames = T,  
  filter = 'top', 
  editable = F, 
  extensions = 'Buttons', 
  options = list(
    pageLength = 6, 
    dom = 'Bfrtip',
    buttons = c('csv')
  ))
```

##  {.unnumbered .unlisted}

$$\\[0.5in]$$

## QC {.tabset .tabset-pills}

### Input reads
```{r}
p <- figs$reads
plotly::ggplotly(p, tooltip = 'text')
```


### Read alignment
```{r}
p <- figs$alignment
plotly::ggplotly(p, tooltip = 'text')

```


### Gene detected and intergenic rate
```{r}
p <- figs$genes
plotly::ggplotly(p, tooltip = 'text')
```

### Exonic vs intronic rates
```{r}
p <- figs$exon
plotly::ggplotly(p, tooltip = 'text')
```


### Metric table
Click a button to save as csv or Excel file. \
```{r}
if (exists('ms')){
  v <- unlist(strsplit('Sample,fastqc.total_sequences,fastqc_trimmed.total_sequences,star.uniquely_mapped,star.uniquely_mapped_percent,PCT_OFF_BAIT, ZERO_CVG_TARGETS_PCT,Genes.Detected,Exonic.Rate,Intronic.Rate,Intergenic.Rate', split = ','))
  colv <- intersect(v, colnames(ms)) 
  df <- ms[,colv]
  colnames(df) <- plyr::mapvalues(colnames(df), from = v, to = c('Sample', 'Total sequenced reads', 'Adpater-trimmed reads', 'Uniquely mapped reads', 'Unique mapping percent', 'Off-target rate', 'Zero coverage rate', 'Genes detected', 'Exonic rate', 'Intronic rate', 'Intergenic rate'))
  
  df %>%
    DT::datatable(
      class = 'cell-border stripe', 
      rownames = T, 
      filter = 'top', 
      editable = T, 
      extensions = 'Buttons', 
      options = list(
        pageLength = 10, 
        dom = 'Bfrtip',
        buttons = c('csv', 'excel')
      ))
}

```

##  {.unnumbered .unlisted}

$$\\[0.5in]$$





## PCA plot of all samples {.tabset .tabset-pills}
See the sample name by hovering over a dot \

### using gene-level expression
```{r, fig.dim=c(6,4)}
library(plotly)
if(exists('gene.pca')){
  ggplotly(gene.pca, tooltip = "text")
  
}


```

### using transcript-level expression
```{r, fig.dim=c(6,4)}
library(plotly)
if(exists('transcript.pca')){
  ggplotly(transcript.pca, tooltip = "text")
  
}



```

##  {.unnumbered .unlisted}

$$\\[0.5in]$$

## Differential gene expression

### DE Summary
feature = gene; FDR = BH-corrected p-value; FC = fold change  \
Note: \
Statistics of DE genes are summarized based on two thresholds -- the first set between columns *features.up* and *FC.cutoff* and the second set between columns *features.up2* and *FC.cutoff2*. \

```{r}
if (exists('de_sum')){
  de_sum %>% 
    relocate(control.samples, test.samples, .after = last_col()) %>% 
    DT::datatable(
      class = 'cell-border stripe', 
      rownames = T,  
      filter = 'top', 
      editable = F, 
      extensions = 'Buttons', 
      options = list(
        pageLength = 6, 
        dom = 'Bfrtip',
        buttons = c('csv')
      ))
}
```

$$\\[0.2in]$$

### DE plots {.tabset .tabset-pills}

DE genes displayed in the plots below and in the delivered pdf files  were defined by the first threshold in section *DE Summary* (between columns *features.up* and *FC.cutoff*) .  \
Note: \
PCA is based on top 500 variable genes (NOT on chr X, chr Y or chr M). \
The PCA plot of samples relevant to a specific comparison are also shown below. \
Comparisons are named as **[test group]\_vs\_[control group]**. \
\

```{r, eval=TRUE, include=TRUE, results='asis', fig.cap=c('Fig. Differential expression analysis. (A) PCA plot. (B) MA/MD plot. (C) Volcano plot'), fig.width=12, fig.height=4}
if (exists('de_plots')){

  for (i in 1:length(de_plots)){
    p <- wrap_plots(de_plots[[i]], nrow = 1) + plot_annotation(tag_levels = 'A')
    
    make_tabs(
      paste("####", names(de_plots)[i]),
      list(p)
    )
  }
  
}

```


$$\\[0.5in]$$



## Differential transcript expression

### DT Summary
feature = transcript; FDR = BH-corrected p-value; FC = fold change  \
Note: \
Statistics of DT genes are summarized based on two thresholds -- the first set between columns *features.up* and *FC.cutoff* and the second set between columns *features.up2* and *FC.cutoff2*. \

```{r}
if (exists('dt_sum')){
  dt_sum %>% 
    relocate(control.samples, test.samples, .after = last_col()) %>% 
    DT::datatable(
      class = 'cell-border stripe', 
      rownames = T,  
      filter = 'top', 
      editable = F, 
      extensions = 'Buttons', 
      options = list(
        pageLength = 6, 
        dom = 'Bfrtip',
        buttons = c('csv')
      ))
}
```

$$\\[0.2in]$$

### DT plots {.tabset .tabset-pills}

Differential transcripts displayed in the plots below and in the delivered pdf files  were defined by the first threshold in section *DT Summary* (between columns *features.up* and *FC.cutoff*) .  \
Note: \
PCA is based on top 500 variable genes (NOT on chr X, chr Y or chr M). \
The PCA plot of samples relevant to a specific comparison are also shown below. \
Comparisons are named as **[test group]\_vs\_[control group]**. \
\

```{r, eval=TRUE, include=TRUE, results='asis', fig.cap=c('Fig. Differential expression analysis. (A) PCA plot. (B) MA/MD plot. (C) Volcano plot'), fig.width=12, fig.height=4}
if (exists('dt_plots')){

  for (i in 1:length(dt_plots)){
    p <- wrap_plots(dt_plots[[i]], nrow = 1) + plot_annotation(tag_levels = 'A')
    
    make_tabs(
      paste("####", names(dt_plots)[i]),
      list(p)
    )
  }
  
}

```





##  {.unnumbered .unlisted}

$$\\[0.5in]$$


## Delivered files
**The following files will be delivered if relevant analyses are performed. ** \

### Arriba/

Genomic alignment bam files, bam index files and detected fusions.  _work/ contains other output files by Arriba.

### {.unnumbered .unlisted}

$$\\[0.1in]$$


### differential_genes/
Differential expression at the gene level. Each comparison has a subfolder (see section *DE summary* for more details). The text file has edgeR output (logFC, logCPM, p-value, BH-adjusted FDR), normalized expression in RPKM and CPM. In column is.sig (and is.sig2) of the text file, each gene is assigned 1 for up-regulation, -1 for down-regulation and 0 for no significant difference based on FDR and log2FC cutoffs (see "**DE summary**" for cutoffs used). PCA, MD/MA and volcano plots are in the same subfolder and also included in this report under section *DE plots*.\


### {.unnumbered .unlisted}

$$\\[0.1in]$$

### differential_transcripts/
Differential expression at the transcript level. Each comparison has a subfolder (see section *DE summary* for more details). The text file has edgeR output (logFC, logCPM, p-value, BH-adjusted FDR), normalized expression in RPKM and CPM. In column is.sig (and is.sig2) of the text file, each transcript is assigned 1 for up-regulation, -1 for down-regulation and 0 for no significant difference based on FDR and log2FC cutoffs (see "**DE summary**" for cutoffs used). PCA, MD/MA and volcano plots are in the same subfolder and also included in this report under section *DE plots*.\


### {.unnumbered .unlisted}

$$\\[0.1in]$$

### expression_quantification/
*all_samples.gene_raw_counts.txt* - Raw gene-level counts of all genes in all samples. \
*all_samples.transcript_raw_counts.txt* - Raw transcript-level counts of all genes in all samples. \

### {.unnumbered .unlisted}

$$\\[0.1in]$$

### MultiQC/
multiqc_report.html is a QC summary report generated using [MultiQC](https://multiqc.info). \

### {.unnumbered .unlisted}

$$\\[0.1in]$$

### QC/

QC by [FastQC](https://www.bioinformatics.babraham.ac.uk/projects/fastqc/) for raw fastq files and adapter-trimmed fastq files, [RSeQC](http://doi.org/10.1093/bioinformatics/bts356) and [RNA-SeQC](https://pubmed.ncbi.nlm.nih.gov/22539670/). \

### {.unnumbered .unlisted}

$$\\[0.5in]$$

### Salmon/
Transcript-level analysis by Salmon. \

### {.unnumbered .unlisted}

$$\\[0.1in]$$

### STAR/

Genomic alignment bam files and bam index files.  _work/ contains other output files by STAR.

### {.unnumbered .unlisted}

$$\\[0.1in]$$

## Methods

[Cutadapt](https://journal.embnet.org/index.php/embnetjournal/article/view/200) was applied to remove adapter sequences. [STAR](https://pubmed.ncbi.nlm.nih.gov/23104886/) with 2-pass mapping was used to align reads to the reference genome and gene annotation along with quantification of gene-level expression. [FastQC](https://www.bioinformatics.babraham.ac.uk/projects/fastqc/), [RNA-SeQC](https://pubmed.ncbi.nlm.nih.gov/22539670/), [RSeQC](https://pubmed.ncbi.nlm.nih.gov/22743226/) were used to check various QC metrics including insert fragment size, read quality, read duplication rates, rRNA rates, gene body coverage and read distribution in different genomic regions. \
Bioconductor package [edgeR](https://academic.oup.com/bioinformatics/article/26/1/139/182458) was used to detect differential expression between sample groups. Genes with low expression were excluded using edgeR function filterByExpr with min.count = 10 and min.total.count = 15. The filtered expression matrix was normalized by [TMM method](https://doi.org/10.1186/gb-2010-11-3-r25). Statistical test was using quasi-likelihood pipeline implemented in edgeR. A gene was deemed differentially expressed if meeting the cutoffs of log2 fold change and Benjamini-Hochberg adjusted p-values. \

$$\\[0.5in]$$

## Acknowledgement
(For publication, please include the following acknowledgement.)
This research was supported by the Genomics & Bioinformatics Shared Resource, RRID:SCR_022606, of the Fred Hutch/University of Washington/Seattle Children’s Cancer Consortium (P30 CA015704).

$$\\[0.5in]$$

